Parallel Training of Deep Neural Network using Model Averaging and Butterfly Mixing

### 2015-06-17 paper.tex
refine the introduction

% Title.
% ------
\title{Investigation on Parallel Training of Deep Neural Network using Model Averaging}
%
% Single address.
% ---------------
\name{Hang Su$^{1,2}$, Haoyu Chen$^1$, Haihua Xu$^3$}
\address{$^1$ International Computer Science Institute, Berkeley, California, US \\
$^2$ Dept. of Electrical Engineering \& Computer Science, University of California, Berkeley, CA, USA \\
$^3$ Nanyang Technological University, Singapore \\
{\small \tt \{suhang3240@gmail.com\}}
}
%
% For example:
% ------------
%\address{School\\
%	Department\\
%	Address}
%
% Two addresses (uncomment and modify for two-address case).
% ----------------------------------------------------------
%\twoauthors
%  {A. Author-one, B. Author-two\sthanks{Thanks to XYZ agency for funding.}}
%	{School A-B\\
%	Department A-B\\
%	Address A-B}
%  {C. Author-three, D. Author-four\sthanks{The fourth author performed the work
%	while at ...}}
%	{School C-D\\
%	Department C-D\\
%	Address C-D}
%
\begin{document}
%\ninept
%
\maketitle
%
\begin{abstract}
In this paper we apply model averaging and butterfly mixing to parallel training of deep neural network (DNN). 
Parallelization is done in a model averaging manner. Data is partitioned and distributed to different nodes for 
local model updates, and model averaging (reduce) is done every few minibatches.

We use multiple GPUs for data parallelization, and Message Passing Interface (MPI) for communication between nodes. 
We investigate the effectiveness of Natrual Gradient for Stochasitc Gradient Descent (NG-SGD) in the model averaging
framework, and compare two learning rate schedule strategies for model update. We show that combination of 
NG-SGD and exponential learning rate schedule can lead to better convergence. On the 300h Swithboard dataset, a 10 
times speed up is achieved using 16 gpus with little decoding accuracy loss -- 0.2 in word error rate (WER). We also 
introduce a butterfly mixing strategy that can further save communication costs in model averaging framework.

\end{abstract}
%
\begin{keywords}
Parallel training, model averaging, deep neural network, butterfly mixing
\end{keywords}
%
\section{Introduction}
\label{sec:intro}
Deep Neural Networks (DNN) has shown its effeciveness in several machine learning tasks, espencially in speech
recognition. The large model size and massive training examples make DNN a powerful model for classification. However,
these two factors also slow down the training procedure of DNNs.

Parallelization of DNN training has been a popular topic since the revive of neural networks. Several different strategies
have been proposed to tackle this problem. Multiple thread CPU parallelization and single CUDA GPU implementation are compared
in \cite{scanzio2010parallel,vesely2010parallel}, and they show that single GPU could beat 8 cores CPU by a factor of 2.

Optimality for parallelization of DNN training was analyzed in \cite{seide2014parallelizability}, and based on the analysis, 
a gradient quantization approach was proposed to minimize communication cost \cite{seide20141}.

DistBelief proposed in \cite{dean2012large} reports a speed up of 2.2x using 8 CPU coresthan using a
single machine.

Asynchronous SGD using multiple GPUs achieved a 3.2x speed-up on 4 GPUs \cite{zhang2013asynchronous}.

@@

A speedup of 6x to 14x was achieved using 16 GPUs on training convolutional neural networks \cite{coates2013deep}. In this approach,
each GPU is responsible for a partition of the neural network. This approach is more useful for image classification where 
local structure of the neural network could be exploited.

Distributed model averaging using CPUs is proposed in \cite{zhang2014improving}, and a further improvement 
is done using natural gradient \cite{povey2014parallel}. Our approach is mainly based on this idea, and the main 
difference between our work and this one is that we are performing neural networks training via MPI. Also,
we introduce butterfly-mixing strategy for model averaging.

Butterfly mixing was proposed in \cite{zhao2013butterfly} to interleave communication with computation. In each
reduce period, information (gradient or model) is exchanged between each node and its friend node. The pair of nodes
are shuffled after every reduce so that information in each node can propagate to all other nodes in 
$log(numNodes)$ periods.

In Section 2 we introduce the model approach. In Section 3 we introduce different reduce strategies. Section 4 
explains NG-SGD. Section 5 records experimental results on different setups and Section 6 concludes.

@@

\subsection{Butterfly mixing}
Butterfly mixing is a reduction strategy proposed in \cite{zhao2013butterfly}. It reduces the communication load in 
every iteration: one node would only send and receive message from one other node. The pairing of nodes are shuffled after
each reduction so that information collected in each nodes will spread to all other nodes. Fig~\ref{fig:butterfly} 
is an example of butterfly mixing with 4 nodes. Butterfly mixing requires much less communication than all-reduce 
strategy. However, each node in the network only collects information from two computing nodes at a time.
It takes $log(numNodes)$ communication to spread the message to the whole network. Thus, the converge speed 
of butterfly mixing would be slower than all-reduce strategy. 
\begin{figure}[htb]
  \centering
  \includegraphics[width=0.5\textwidth]{butterfly.png}
  \caption{Butterfly mixing network}
  \label{fig:butterfly}
\end{figure}

>>>>

% Title.
% ------
\title{Parallel Training of Deep Neural Network using Model Averaging and Butterfly Mixing}
%
% Single address.
% ---------------
\name{Hang Su$^{1,2}$, Haoyu Chen$^1$, Haihua Xu$^3$}
\address{$^1$ International Computer Science Institute, Berkeley, California, US \\
$^2$ Dept. of Electrical Engineering \& Computer Science, University of California, Berkeley, CA, USA \\
$^3$ Nanyang Technological University, Singapore \\
{\small \tt \{suhang3240@gmail.com\}}
}
%
% For example:
% ------------
%\address{School\\
%	Department\\
%	Address}
%
% Two addresses (uncomment and modify for two-address case).
% ----------------------------------------------------------
%\twoauthors
%  {A. Author-one, B. Author-two\sthanks{Thanks to XYZ agency for funding.}}
%	{School A-B\\
%	Department A-B\\
%	Address A-B}
%  {C. Author-three, D. Author-four\sthanks{The fourth author performed the work
%	while at ...}}
%	{School C-D\\
%	Department C-D\\
%	Address C-D}
%
\begin{document}
%\ninept
%
\maketitle
%
\begin{abstract}
In this paper we apply model averaging and butterfly mixing to parallel training of deep neural network (DNN). 
Parallelization is done in a model averaging manner. Data is partitioned and distributed to different nodes for 
local model updates, and model averaging (reduce) is done every few minibatches.

We use multiple GPUs for data parallelization, and Message Passing Interface (MPI) for communication between nodes. 
We investigate the effectiveness of Natrual Gradient for Stochasitc Gradient Descent (NG-SGD) in the model averaging
framework, and compare two learning rate schedule strategies for model update. We show that combination of 
NG-SGD and exponential learning rate schedule can lead to better convergence. On the 300h Swithboard dataset, a 10 
times speed up is achieved using 16 gpus with little decoding accuracy loss -- a 0.2\% degradation in word error rate 
(WER). We also introduce a butterfly mixing strategy that can further save communication costs in model averaging framework.
\end{abstract}
%
\begin{keywords}
Parallel training, model averaging, deep neural network, butterfly mixing
\end{keywords}
%
\section{Introduction}
\label{sec:intro}
Deep Neural Networks (DNN) has shown its effeciveness in several machine learning tasks, espencially in speech
recognition. The large model size and massive training examples make DNN a powerful model for classification. However,
these two factors also slow down the training procedure of DNNs.

Parallelization of DNN training has been a popular topic since the revival of neural networks. Several different strategies
have been proposed to tackle this problem. Multiple thread CPU parallelization and single GPU implementation are compared
in \cite{scanzio2010parallel,vesely2010parallel}, and it is shown that single GPU could beat multi-threaded CPU implementation
by a factor of 2.

Optimality for parallelization of DNN training was analyzed in \cite{seide2014parallelizability}, and based on the analysis, 
a gradient quantization approach (1-bit SGD) was proposed to minimize communication cost \cite{seide20141}. It shows that 1 bit
can effectively reduce the amount of data exchanged in an MPI framework, and a 10 times speed-up using 40 GPUs.

DistBelief proposed in \cite{dean2012large} reports that 8 CPU machines train 2.2 times faster than a single GPU machine on a
moderately sized speech model. Asynchronous SGD using multiple GPUs achieved a 3.2x speed-up on 4 GPUs \cite{zhang2013asynchronous}.

@@

A speedup of 6x to 14x was achieved using 16 GPUs on training convolutional neural networks \cite{coates2013deep}. In this approach,
each GPU is responsible for a partition of the neural network. This approach is more useful for image classification where 
local structure of the neural network could be exploited. For a fully connected speech model, a model partition approach 
may not be able to contribute as much.

Distributed model averaging using CPUs is proposed in \cite{zhang2014improving}, and a further improvement 
is done using natural gradient \cite{povey2014parallel}. In this approach, separate models are trained on multiple nodes using 
different partitions of data, and model parameters are averaged after each epoch. It is shown that NG-SGD can effectively improve
convergence and ensure a better model trained using the model averaging framework.

Our approach is mainly based on the NG-SGD with model averaging. The main differences here in this work are:
1. We utilize multiple GPUs in neural networks training via MPI, which allows us to perform model averaging more frequently and efficiently;
2. We introduce a butterfly-mixing strategy for model averaging that can further speed up the training.

In addition, this work simplifies the model training procedure in the following aspects:
1. It does not require a warm-up phase where only single thread can be used for model update \cite{seide20141,povey2014parallel};
2. It does not need to tune minibatch size to utilize parallelization power;
3. It does not use momentum or Adagrad\cite{duchi2011adaptive} for gradient estimation.

On the other hand, it requires:
1. A proper implementation of NG-SGD, as is introduced in \cite{povey2014parallel};
2. A bigger initial learning rate and a proper model averaging frequency;

In Section 2 we introduce the model averaging approach. In Section 3 we introduce butterfly-mixing strategie. Section 4 
explains NG-SGD. Section 5 records experimental results on different setups and Section 6 concludes.

@@

\subsection{Butterfly mixing}
Butterfly mixing was proposed in \cite{zhao2013butterfly} to interleave communication with computation. In each
reduce period, information (gradient / model) is exchanged between each node and its friend node. The pair of nodes
are shuffled after every reduce so that information in each node can propagate to all other nodes in $log(numNodes)$ periods.

Butterfly mixing is a reduction strategy proposed in \cite{zhao2013butterfly}. It reduces the communication load in 
every iteration: one node would only send and receive message from one other node. The pairing of nodes are shuffled after
each reduction so that information collected in each nodes will spread to all other nodes. Fig~\ref{fig:butterfly} 
is an example of butterfly mixing with 4 nodes. Butterfly mixing requires much less communication than all-reduce 
strategy. However, each node in the network only collects information from two computing nodes at a time.
It takes $log(numNodes)$ communication to spread the message to the whole network. Thus, the converge speed 
of butterfly mixing would be slower than all-reduce strategy. 
\begin{figure}[htb]
  \centering
  \includegraphics[width=0.5\textwidth]{butterfly.png}
  \caption{Butterfly mixing network}
  \label{fig:butterfly}
\end{figure}
